What is an Ancestral Recombination Graph?

### 2022-02-18 paper.tex
more text

% What do we mean by an EARG, formally?
% NOTE we're using the natural numbers but are using 0 as a node in the
% examples. Probably want to clear this up a bit.
Formally, we can define an EARG as a tuple $(e, \sigma)$, where $e$ is an ordered
list of tuples such that $e_j = (c, p)$ defines a directed edge between
child node $c$ and parent $p$ ($c, p \in \mathbb{N}$), and
$\sigma: \mathbb{N} \rightarrow \mathbb{N}$
is a function mapping nodes to recombination breakpoints, such that $\sigma(u) = x$
if $u$ is a recombination event with breakpoint $x$ and
$\sigma(u) = \infty$ otherwise. (The use of $\infty$ to denote
non-recombination nodes is arbitrary, but useful when we extract
% NOTE: need to be careful about using "local" trees if we're also
% distinguishing resolved and non-resolved GARGs as being "locally"
% defined or not. One is topologically local and the other is local
% along the genome.
local trees; see below.)
For each recombination node $u$ we must
have exactly two edges in $E$ such that $u$ is the child.
Note that this description captures only the
graph topology: if we also wish to know the times of events we need
an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
which defines the time of each node.
An example is given in Fig.~\ref{fig-arg-data-structure}A.

@@

We may have nodes corresponding to genomes that are ancestral to
the sample along their entire span
(e.g.\ node $H$ in Fig.~\ref{fig-ancestry-resolution}),
genomes that carry no genetic material ancestral to the sample at all
(e.g.\ node $J$ in Fig.~\ref{fig-ancestry-resolution}),
and every possibility along the continuum between these extremes.
Clearly, we can only ever hope to infer the fraction of an ancestor's
genome that is ancestral to the sample.

\begin{figure}
\centering
\vspace{5em}
\includegraphics[width=\linewidth]{illustrations/ARG_recomb_node_deletion}
\caption{\label{fig-simplification}
ARG simplification. (a) A simulated
ARG including all recombination nodes (red). (b) The same ARG having removed both
recombination nodes and nodes of common ancestors in which no genetic
lineages coalesce (cyan). (c) The sequence of trees corresponding to (b): if we were
to plot out the trees corresponding to (a) we would have more trees, but the topologies
and MRCA nodes at each genomic location would be identical. For clarity, nodes are
plotted by ranked time rather than true time on the y-axis.
% HYW: note that we also truncate nodes for portions in which they are unary,
% so that nodes which are coalescent nodes over only some of their length
% (blue-green) become coalescent nodes over their entire length (green). This
% can, however, throw away some useful information, for example node 4 has
% only a single parent edge in (a), but 2 parent edges (one of which appears
% to bypass node 11) in (b). In fact, this edge should go through node 11, but
% there may be no way to detect this from real data.
}
\end{figure}

>>>>

% What do we mean by an EARG, formally?
% NOTE we're using the natural numbers but are using 0 as a node in the
% examples. Probably want to clear this up a bit.
Formally, we can define an EARG as a tuple $(e, \sigma)$, where $e$
defines a directed acyclic graph where $e_j = (c, p)$ describes
an edge between
child node $c$ and parent $p$ ($c, p \in \mathbb{N}$).
Additionally,
$\sigma: \mathbb{N} \rightarrow \mathbb{N}$
is a function mapping nodes to recombination breakpoints, such that $\sigma(u) = x$
if $u$ is a recombination event with breakpoint $x$ and
$\sigma(u) = \infty$ otherwise. (The use of $\infty$ to denote
non-recombination nodes is arbitrary, but useful when we extract
% NOTE: need to be careful about using "local" trees if we're also
% distinguishing resolved and non-resolved GARGs as being "locally"
% defined or not. One is topologically local and the other is local
% along the genome.
local trees; see below.)
For each recombination node $u$ we must
have exactly two edges in $E$ such that $u$ is the child.
Note that this description captures only the
graph topology: if we also wish to know the times of events we need
an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
which defines the time of each node.
An example is given in Fig.~\ref{fig-arg-data-structure}A.

@@

\begin{figure}
\centering
\vspace{5em}
\includegraphics[width=\linewidth]{illustrations/ARG_recomb_node_deletion}
\caption{\label{fig-simplification}
ARG simplification. (A) An ARG simulated from a diploid Wright-Fisher
model. Note the complex full-sib relationship between B and C; this
could not be directly represented in a Griffiths-like EARG encoding
(see section XXX).
(B) Simplified to remove all
% trying out this terminology - can define in the text
1-connected graph components (e.g., diamonds).
(C) Remove nodes that are unary in all local trees.
(D) Rewrite edges to skip any unary nodes in local trees.
(E) The local trees for ARG (A).
(F) The local trees for ARG (D).
}
\end{figure}

% Q: would it help to talk about "tree vertices" here or would it
% just be more confusing? I worry that people would think that
% they are two different things then, and it's really important
% that they get that a node-is-a-node
The ideas of ``meaningful'' local tree topologies revolve around the
presence of ``unary'' nodes: a node in the tree that has exactly
one child. For example, Fig.~\ref{fig-simplification}E shows the
local trees along the genome for a simulated ARG
(Fig.~\ref{fig-simplification}A). Consider the right-most tree,
and the path above node $D$. Before $D$ meets its common ancestor
over the interval $(93, 100]$ with the other samples it passes through
nodes $H, K, N, R$ and $S$, reflecting the corresponding
traversal through the ARG (see section XXX). Such unary nodes
in a local tree are not usually considered significant,
because they provide very little information and are not
detectable by information from the tree. % weak
The nodes in a tree usually mark the points where samples find a
\emph{common} ancestor, rather than focusing on specific
ancestors. For most purposes, we would consider an edge
directly joining sample $D$ to the node $T$ to be more
useful and informative. Simplification is the process of removing
these uninformative and (for most computational purposes)
inefficient unary paths.

The first type of simplification that we can perform is to remove
graph topology that is invisible to the samples. The most well
known example of such topology is a so-called
``diamond''~\citep{rasmussen2014genome}
in which the two parent nodes of a recombination immediately
join again into a common ancestor (e.g. $K,N,O$ and $R$ in
Fig.~\ref{fig-simplification}A).
Unless we are specifically
interested in the recombination event or ancestral genomes,
there is no information in this topology and the diamond can b
e replaced by a single edge. More generally, any
subgraph which is singly-connected in both the leafward and
rootward direction (a ``super-diamond'') is non-identifiable and can be
replaced by a single edge. This definition includes the case
of a single node that has one inbound and one outbound edge.
Fig.~\ref{fig-simplification}B shows the result of this type of
graph-level simplification.

Simplifying away diamonds will remove many unary nodes from the
local trees, there can still be nodes that are unary in all
of the local trees. For example, node $H$ is one of the
parents of the recombinant node $D$ is still present in
Fig.~\ref{fig-simplification}B after diamond removal, but
it is unary in all of the local trees since it has an
in-degree of $1$.
Fig.~\ref{fig-simplification}C shows the resulting graph
after removing these nodes.

The remaining nodes are MRCAs of some subset of the samples
at \emph{some} positions along the genome. We still have
some unary nodes in the local trees, but these nodes will
correspond to a coalescence in at least one other
local tree. For example, node $E$ is unary in the first tree
of Fig.~\ref{fig-simplification}E, but is either binary
or ternary in the other trees (recall this is a Wright-Fisher
simulation).

The final level of simplification is to remove \emph{all}
unary nodes from the local trees (Fig.~\ref{fig-simplification}D
and E). [And continue]

% This text might go somewhere
% We may have nodes corresponding to genomes that are ancestral to
% the sample along their entire span
% (e.g.\ node $H$ in Fig.~\ref{fig-ancestry-resolution}),
% genomes that carry no genetic material ancestral to the sample at all
% (e.g.\ node $J$ in Fig.~\ref{fig-ancestry-resolution}),
% and every possibility along the continuum between these extremes.
% Clearly, we can only ever hope to infer the fraction of an ancestor's
% genome that is ancestral to the sample.
